---
title: "Demonstration: Quarto, graphs, tables, and websites"
author: "Rohan Alexander"
date: "2026-09-16"
date-format: "D MMMM, YYYY"
format: pdf
bibliography: references.bib
---

# What Quarto is

## What even is Quarto?

### What, seriously, is Quarto?

Quarto turns a plain-text file with a `.qmd` extension into a PDF, a Word document, a slide deck, or a website. The file has three kinds of content: a YAML header between `---` lines that says what to produce; Markdown prose; and code chunks that run when the file renders, so the numbers, graphs, and tables in the output come from the data every time. This document is a `.qmd`; open `02-demo.qmd` next to the PDF and compare.

A paper where the figures were pasted in by hand cannot be checked and cannot be updated when the data change. A Quarto paper can be rendered by anyone with the repo.

# The workflow from last week

## Step 1: Python makes the data

`aggregate_shelter.py` downloads the City of Toronto's daily shelter occupancy file (the Week 1 data), keeps the columns we need, and aggregates to one row per date and sector. Run it once:

```bash
uv run aggregate_shelter.py
```

It writes `data/shelter_daily.parquet`. Parquet rather than CSV!

## Step 2: R reads it

This chunk runs when this document renders. The first one reads the Parquet file with the `arrow` package.

```{r}
#| message: false
#| warning: false

library(arrow)
library(dplyr)
library(ggplot2)
library(tinytable)

daily <- read_parquet("data/shelter_daily.parquet")

glimpse(daily)
```

## Step 3: a graph


```{r}
#| label: fig-users
#| fig-cap: "People in Toronto's shelter system each night in 2026, by sector. Data from Open Data Toronto."
#| fig-width: 8
#| fig-height: 4

daily |>
  ggplot(aes(x = date, y = service_users, colour = sector)) +
  geom_line() +
  labs(x = "Date", y = "People", colour = "Sector") +
  theme_classic()
```

@fig-users is a cross-reference: the `label` on the chunk starts with `fig-`, so Quarto numbers the figure and the reference updates itself if figures move. Three things: the graph has labelled axes with units, the legend is named, and the ordering of the lines carries information.

## Step 4: a table

```{r}
#| label: tbl-latest
#| tbl-cap: "Shelter occupancy on the most recent night in the data, by sector."

latest <- 
  daily |>
  filter(date == max(date)) |>
  mutate(
    beds = if_else(capacity_beds > 0, occupied_beds / capacity_beds, NA),
    rooms = if_else(capacity_rooms > 0, occupied_rooms / capacity_rooms, NA)
  ) |>
  select(
    Sector = sector, People = service_users,
    `Bed occupancy` = beds, `Room occupancy` = rooms
  )

tt(latest, digits = 3) |>
  format_tt(
    j = c("Bed occupancy", "Room occupancy"),
    fn = scales::label_percent(accuracy = 0.1)
  ) |>
  format_tt(escape = TRUE)
```

@tbl-latest is the same idea for a table. The table has a caption that says what the rows are and when the data are from. In terms of measurement, notice that Families may be room-based and so could undercount.

## Step 5: Citations

Citations come from `references.bib` and the `bibliography` line in the header. Writing `@healy2019` gives @healy2019, and `[@healy2019]` gives a parenthetical [@healy2019]. The reference list is added at the end automatically.

# Rendering

From the terminal, in the folder that has the file:

```bash
quarto render 02-demo.qmd
```

`format: pdf` in the header makes a PDF (this needs a LaTeX installation; `quarto install tinytex` gets one). Change it to `format: html` for a web page or `format: revealjs` for slides; nothing else in the file changes. In VS Code, the Quarto extension gives you a Render button and a preview.

# Websites

A Quarto website is a folder with a `_quarto.yml` and some `.qmd` pages. `website_example/` in this week's folder is a working one; copy it. The three files that matter:

`_quarto.yml`:

```yaml
project:
  type: website
  output-dir: docs

website:
  title: "Toronto shelters"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: shelter.qmd
        text: Occupancy

format:
  html:
    theme: cosmo
```

`index.qmd` is the home page, `shelter.qmd` is a page with the graph and table above. Rendering the whole site:

```bash
quarto render
```

puts the finished HTML in `docs/`. That folder name is the trick for GitHub Pages: push the repo to GitHub, then under Settings, Pages, choose "Deploy from a branch", branch `main`, folder `/docs`. A minute later the site is at `https://<username>.github.io/<repo>/`. Every time you render and push, the site updates. Commit `docs/`; it is the product.

# Working in groups: pull requests and conflicts

Next week's ISLR/P group needs to get its slides into the class repo, and the Part 2 teams will share one repo for a month. The mechanism for both is the pull request.

The loop, for a repo you do not own:

```bash
# once: fork on GitHub, then
git clone https://github.com/<you>/<repo>.git
cd <repo>
git checkout -b week_03_group_1        # a branch for your change
# ... add your files ...
git add .
git commit -m "Add group 1 slides on ISLR 2.1"
git push -u origin week_03_group_1
```

Then on GitHub, "Compare & pull request". The PR is where the review happens: someone else reads the diff, leaves comments on lines, and approves or asks for changes. Only after that does it get merged. Nobody pushes to `main` directly.

A conflict happens when two people change the same lines and the second one to merge has to choose. Git marks the file like this:

```
<<<<<<< HEAD
- Group 1: Chapter 2.1
=======
- Group 1: Chapter 5.1
>>>>>>> week_03_group_1
```

Open the file, keep the version you want, delete the three marker lines, then `git add` and `git commit`. That is all a conflict is. The Timbers reading has the longer version; the live demonstration will make one on purpose with two branches that edit the same line of a README.

Two rules that prevent most conflicts: pull before you start (`git pull`), and keep PRs small and about one thing.

# What to do in the worksheet

Question 3 builds a website like `website_example/` and puts it on GitHub Pages. Question 2 is your first PR to the class repo. Question 4 is three graphs and a table of your own, on the website. If you finish, make a conflict with your neighbour and fix it.

\newpage

# References
