---
title: "ISLR 3.1–3.2: 10-minute Linear Regression Lab"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Goal

Use the `Advertising` data to reproduce the central ideas from Sections 3.1–3.2:

- simple vs multiple linear regression;
- coefficient interpretation;
- $R^2$ and residual error;
- how adjustment changes the newspaper coefficient;
- confidence vs prediction intervals.

## Setup

```{r}
library(tidyverse)

Advertising <- read_csv(
  "https://www.statlearning.com/s/Advertising.csv"
)

Advertising
```

## 1. Simple linear regression

Fit sales on TV advertising.

```{r}
fit_tv <- lm(sales ~ TV, data = Advertising)
summary(fit_tv)
```

**Write down:**

1. The fitted equation $\widehat{sales}=\hat\beta_0+\hat\beta_1TV$.
2. Interpret $\hat\beta_1$ in the units of this data.
3. What proportion of variation in sales is explained by TV alone?
4. What is the estimated standard deviation for the error term?

```{r}
coef(fit_tv)
summary(fit_tv)$r.squared
sigma(fit_tv) # RSE
```

## 2. Visualize residuals

```{r}
plot(Advertising$TV, Advertising$sales,
     xlab = "TV advertising ($1000s)",
     ylab = "Sales (1000s of units)")
abline(fit_tv, lwd = 2)
```

Pick one observation mentally. Its vertical distance from the fitted line is its **residual**.

**Question:** Does a straight line appear to capture the main trend?

## 3. Newspaper by itself

```{r}
fit_news <- lm(sales ~ newspaper, data = Advertising)
summary(fit_news)
```

Record the newspaper coefficient and p-value.

## 4. Multiple regression

```{r}
fit_all <- lm(sales ~ TV + radio + newspaper, data = Advertising)
summary(fit_all)
```

**Questions:**

1. Interpret the radio coefficient using the phrase **“holding TV and newspaper fixed.”**
2. What happened to the newspaper coefficient and p-value?
3. Compare $R^2$ and RSE with the TV-only model.

```{r}
c(
  TV_only_R2 = summary(fit_tv)$r.squared,
  all_R2 = summary(fit_all)$r.squared,
  TV_only_RSE = sigma(fit_tv),
  all_RSE = sigma(fit_all)
)
```

## 5. Why did newspaper change?

```{r}
cor(Advertising[, c("TV", "radio", "newspaper", "sales")])
```

Look specifically at the correlation between `radio` and `newspaper`.

**Explain in 1–2 sentences:** Why can newspaper be significant in `sales ~ newspaper`, but not in `sales ~ TV + radio + newspaper`?

## 6. Overall F-test

Look again at:

```{r}
summary(fit_all)
```

The F-test asks

$$H_0:\beta_{TV}=\beta_{radio}=\beta_{newspaper}=0$$

against the alternative that **at least one** is non-zero.

**Question:** What does the very small F-test p-value tell us? What does it *not* tell us about each individual predictor?

## 7. Confidence vs prediction interval

Suppose a market has TV = 100, radio = 20, and newspaper = 0.

```{r}
new_market <- data.frame(TV = 100, radio = 20, newspaper = 0)

predict(fit_all, newdata = new_market,
        interval = "confidence", level = 0.95)

predict(fit_all, newdata = new_market,
        interval = "prediction", level = 0.95)
```

**Question:** Which interval is wider, and why?

## If you finish early

Fit the reduced model without newspaper:

```{r}
fit_reduced <- lm(sales ~ TV + radio, data = Advertising)
summary(fit_reduced)
```

Compare it with the full model:

```{r}
anova(fit_reduced, fit_all)
```

Does newspaper add meaningful explanatory power after TV and radio are already in the model?

## Takeaway

The most important conceptual distinction is:

> A simple-regression coefficient describes a marginal association; a multiple-regression coefficient describes an association after holding the other included predictors fixed.
